On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-16 paper/paper3d.tex
regular update

\begin{abstract}
Low mass, self-gravitating accretion disks can remain in a  
quasi-steady, `gravito-turbulent' state provided that thermal 
time-scales are sufficiently long, or equivalently for sufficiently
weak turbulent stresses. Otherwise, numerical simulations show that
disk fragmentation ensues. 
We quantify the interpretation of this result as the
compromise of gravitational stability by non-ideal physics: 
\begin{inparaenum}[1)] 
\item cooling, which reduces pressure support; and
\item viscosity, which reduces rotational support. 
\end{inparaenum}
To this end, we analyse the gravitational stability of accretion disks
including these effects explicitly.   
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and explicit cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to gravito-turbulence, modeled as an effective
viscosity, and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial 
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. We 
thus suggest that disk fragmentation can be physically attributed, 
at least partially, to the removal of rotational stabilization against
self-gravity by turbulence.   
\end{abstract}
%alternative abstract

>>>>

\begin{abstract}
Low mass, self-gravitating accretion disks can remain in a  
quasi-steady, `gravito-turbulent' state provided that thermal 
time-scales are sufficiently long, or equivalently for sufficiently
weak turbulent stresses. Otherwise, numerical simulations show that
disk fragmentation ensues. 
We quantify the interpretation of this result as the
compromise of gravitational stability by: 
\begin{inparaenum}[1)] 
\item 
cooling, which reduces pressure support; and/or
\item 
viscosity, which reduces rotational support. 
\end{inparaenum}
We analyse the gravitational stability of accretion disks
including these non-ideal effects explicitly.   
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to gravito-turbulence, modeled as an effective
viscosity, and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial 
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. We 
thus suggest that disk fragmentation can be physically attributed, 
at least partially, to the removal of rotational stabilization against
self-gravity by turbulence.   
\end{abstract}
%alternative abstract

### 2016-01-16 paper/results3d.tex
regular update

\section{Three-dimensional disks with beta cooling}\label{3ddisk}
We now extend the previous results to 3D disks with vertical
structure. This is expected to weaken gravitational instabilities by
`diluting' self-gravity along the vertical direction. It is possible to incorportate
this effect in the previous 2D framework, but this introduces
additional parameters. We discuss this in Appendix \ref{3dcorr}. 
It is more direct to solve the 3D eigenvalue problem to avoid such 
uncertainties.

{\bf note: fast beta-cooling in 3D destablizes vertical modes (enables
convection)}

%take k>0 wlog
%consider Gamma=1
%height s.t. rho(zmax) = 0.05 rho(0) avoid very low densities 
\subsection{Numerical procedure}
We use a pseudo-spectral method to solve the set of ordinary
differential equations, Eq. \ref{lin_mass}---\ref{lin_gravity}, on 
the domain $ z\in[0,\zmax]$ with a parity condition at the midplane.
We expand $\bm{U}=[\delta P,\delta\rho,\delta\Phi,
  \delta v_x, \delta 
  v_y]$ in even Chebyshev polynomials,
\begin{align}
  \bm{U}(z) = \sum_{j=1}^N \bm{a}_jT_{2(j-1)}(z/\zmax), 
\end{align}
and the vertical velocity in odd Chebyshev polynomials,
\begin{align}
  \delta v_z(z) = \sum_{j=1}^N b_jT_{2j-1}(z/\zmax),   
\end{align}
where $\bm{a}_j$ and $b_j$ are the spectral coefficients. 
The basis functions are chosen to satisfy a reflecting boundary
condition at the midplane, %GI has even symmetry across midplane 
\begin{align}
  \bm{U}^\prime(0) = \bm{0}, \quad \delta v_z(0) = 0.
\end{align}
%At the upper disk boundary we also apply reflection,
We explicitly apply a reflecting upper disk boundary, 
\begin{align}
  \delta v_z(\zmax) = \delta v_x^\prime(\zmax) = \delta
  v_y^\prime(\zmax) = 0,
\end{align}
and the potential satisfies
\begin{align}
  \delta \Phi^\prime (\zmax) + k \delta\Phi(\zmax) = 0, 
\end{align}
as derived by \cite{goldreich65a} and used in similar studies  
\citep{kim12,lin14c}. %reflection is standard in direct simulations

@@

In the following examples we consider a vertically isothermal disk
($\Gamma=1$) with a numerical surface such 
that $\rho(\zmax)=0.05\rho_0$, and use $N=65$.

@@

Fig. \ref{3d_inviscid} show growth rates and the most unstable wavenumber
obtained for a 3D inviscid disk without irradiation. We also plot 2D
results with the 3D correction as described in Appendix
\ref{3dcorr}. While the chosen value of $H_\mathrm{sg}$ results in a close
match between 2D and 3D growth rates, the 3D wavenumber is somewhat
smaller. This reflects self-gravity being weakened in the verical
direction: a larger horizontal scale is required to achieve the same
strength of self-gravity (and hence growth rate) as the 2D case.

@@

Fig. \ref{3d_visc} show growth rates, maximized over $k$, as a
function of the cooling time $\beta$. We also plot 2D results with 3D
corrections. Softening the self-gravity in 2D captures the correct 
qualitative behavior as in the full 3D case, but it is clear that a
single, constant value of $H_\mathrm{sg}$ cannot re-produce 3D growth 
rates for all $\beta$. 
This suggests that the exact value of
$H_\mathrm{sg}$ is problem-dependent (cooling and viscosity), although taking
$H_\mathrm{sg}\sim O(H)$ should give the correct 3D growth rate within
a factor of two. In this respect it is easier to
directly solve the 3D problem.

>>>>

\section{Three-dimensional disks with beta cooling}\label{3ddisk}
We briefly confirm the above results in 3D disks with vertical
structure. This is expected to weaken gravitational instabilities
because the disk mass is spread across the vertical dimension. It is
possible to incorportate this effect in the previous 2D framework, but this
introduces additional parameters. We discuss this in Appendix
\ref{3dcorr}. It is more direct to solve the 3D eigenvalue problem to
avoid such uncertainties. 
%We will find that while softening the
%self-gravity in 2D can produce qualitatively similar results to 3D, a
%quantitative match is

%{\bf note: fast beta-cooling in 3D destablizes vertical modes (enables
%convection)}

%take k>0 wlog
%consider Gamma=1
%height s.t. rho(zmax) = 0.05 rho(0) avoid very low densities 
\subsection{Numerical procedure}
{\bf move to appendix?}
We use a pseudo-spectral method to solve the set of ordinary
differential equations, Eq. \ref{lin_mass}---\ref{lin_gravity}, on 
the domain $ z\in[0,\zmax]$ with a parity condition at the midplane.
We expand $\bm{U}=[\delta P,\delta\rho,\delta\Phi,
  \delta v_x, \delta 
  v_y]$ in even Chebyshev polynomials,
\begin{align}
  \bm{U}(z) = \sum_{j=1}^N \bm{a}_jT_{2(j-1)}(z/\zmax), 
\end{align}
and the vertical velocity in odd Chebyshev polynomials,
\begin{align}
  \delta v_z(z) = \sum_{j=1}^N b_jT_{2j-1}(z/\zmax),   
\end{align}
where $\bm{a}_j$ and $b_j$ are the spectral coefficients. 
The basis functions are chosen to satisfy a reflecting boundary
condition at the midplane, %GI has even symmetry across midplane 
\begin{align}
  \bm{U}^\prime(0) = \bm{0}, \quad \delta v_z(0) = 0.
\end{align}
%At the upper disk boundary we also apply reflection,
We explicitly apply a reflecting upper disk boundary, 
\begin{align}
  \delta v_z(\zmax) = \delta v_x^\prime(\zmax) = \delta
  v_y^\prime(\zmax) = 0,
\end{align}
and the potential satisfies
\begin{align}
  \delta \Phi^\prime (\zmax) + k \delta\Phi(\zmax) = 0, 
\end{align}
as derived by \cite{goldreich65a} and used in similar studies  
\citep{kim12,lin14c}. %reflection is standard in direct simulations

@@

In the following examples we consider a vertically isothermal disk
($\Gamma=1$). Then in the viscous case $\alpha$ is vertically
constant (Eq. \ref{alpha_beta_relation}). We use $N=65$  and apply
numerical disk surface such that $\rho(\zmax)=0.05\rho_0$.

@@

Fig. \ref{3d_inviscid} show growth rates and the most unstable wavenumber
obtained for a 3D inviscid disk with $\qthree=0.71$
($Q=2Q_\mathrm{crit}$) and $\theta=0$. We also plot 2D
results with the 3D correction as described in Appendix
\ref{3dcorr}. The (empirically) chosen value of
$H_\mathrm{sg}=0.64H$ results in a close match between 2D and 3D
growth rates, but the most unstable wavenumber in 3D is somewhat smaller. 
This reflects self-gravity being weakened in the verical 
direction: a larger horizontal scale is required to achieve the same
strength of self-gravity as the 2D case. Similarly, 
choosing $H_\mathrm{sg}=0.53H$ matches the optimum wavenumbers, but
growth rates are over-estimated in 2D.

@@

Fig. \ref{3d_visc} show growth rates, maximized over $k$, as a
function of the cooling time $\beta$. We also plot 2D results with 3D
corrections. Softening the self-gravity in 2D captures the correct 
qualitative behavior as in the full 3D case, but it is clear that a
single, constant value of $H_\mathrm{sg}$ cannot re-produce 3D growth 
rates for all $\beta$. 
This suggests that the exact value of
$H_\mathrm{sg}$ is problem-dependent, although taking
$H_\mathrm{sg}\sim O(H)$ should give the correct 3D growth rate within
a factor of two. 
%A smaller $H_\mathrm{sg}$ is needed with faster 
%cooling (larger viscosity) and vice versa. 
%because most unstable wavenumber is smaller with high viscosity

%smaller scales with increasing beta

%, because viscosity makes the
%disk behave more two-dimensional.

%by eliminating vertical velocity gradients.
